---
title: "CH2"
author: "Mengqi Zhao"
date: "February 11, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
library(lubridate)
library(dplyr)
library(xts)
library(dygraphs)
library(plotly)
```


# CH2 Intro to R & Work with Time Series Date
## Get to know R
### Lesson 1 - Get to know R
```{r eval=FALSE}
# load the ggplot2 library for plotting
library(ggplot2)

# download data from figshare
# note that you are downloading the data into your
download.file(url = "https://ndownloader.figshare.com/files/7010681", destfile = "data/boulder-precip.csv")

# import data
boulder_precip <- read.csv(file="data/boulder-precip.csv")

# view first few rows of the data
head(boulder_precip)

# when you download the data you create a dataframe
# view each column of the data frame using it's name (or header)
boulder_precip$DATE

# view the precip column
boulder_precip$PRECIP

# q plot stands for quick plot. Let's use it to plot your data
qplot(x = boulder_precip$DATE,
      y = boulder_precip$PRECIP)
```

### Lesson 2 & 3 - Variables, Data Types, Vector Data
```{r}
num_char <- c(1, 2, 3, 'a')
num_logical <- c(1, 2, 3, '2.45')
#char_logical <- c('a', 'b', 'c', frog)
tricky <- c(1, 2, 3, '4')
class(num_char)
class(num_logical)
# class(char_logical)
class(tricky)

weight_g <- c(50, 60, 65, 82)
weight_g > 50
weight_g[weight_g > 50]
weight_g[weight_g < 30 | weight_g > 50]
weight_g[weight_g < 30 & weight_g > 50]

animals <- c("mouse", "rat", "dog", "cat")
animals[animals == "cat" | animals == "rat"] # returns both rat and cat
# The function %in% allows you to test if a value is found in a vector
animals %in% c("rat", "cat", "dog", "duck")
animals[animals %in% c("rat", "cat", "dog", "duck")]

```

### Lesson 4 - Import, Work with an dPlot Spreadsheed (Tabular) Data
#### Challenge

* List 3 arguments that are available in the read.csv function.
* How do you figure out what working directory you are in?
* List 2 ways to set the working directory in RStudio.
* Explain what the $ is used for when working with a data.frame in R.
* When you use read.csv are you executing a: a) function or b) variable ?
```{r}
args(read.csv)
getwd()
# setwd() and used the manu
#function

```

### Lesson 5 - How to Address Missing Values
```{r eval=FALSE}
download.file("https://ndownloader.figshare.com/files/9282364", "data/boulder-precip.csv", method = "libcurl")
boulder_precip <- read.csv(file = "data/boulder-precip.csv")
str(boulder_precip)
mean(boulder_precip$PRECIP)
mean(boulder_precip$TEMP)
boulder_precip$TEMP
# plot the data with ggplot
ggplot(data = boulder_precip, aes(x = DATE, y = TEMP)) +
  geom_point() +
  labs(title = "Temperature data for Boulder, CO")

heights <- c(2, 4, 4, NA, 6)
mean(heights)
max(heights)
mean(heights, na.rm = TRUE)
max(heights, na.rm = TRUE)
heights[!is.na(heights)]
na.omit(heights)
heights[complete.cases(heights)]

mean(boulder_precip$TEMP, na.rm = TRUE)
summary(boulder_precip$TEMP, na.rm = TRUE)

boulder_precip_na <- read.csv(file = "data/boulder-precip.csv",
                     na.strings = c("NA", " ", "-999"))
boulder_precip_na$TEMP
ggplot(data = boulder_precip_na, aes(x = DATE, y = TEMP)) +
  geom_point() +
  labs(title = "Temperature data for Boulder, CO",
       subtitle = "missing data accounted for")

sample <- c(2, 4, 4, "NA", 6)
mean(sample, na.rm = TRUE)
```

### Lesson 6 - Plot Data and Customize Plots
```{r}
boulder_precip <- read.csv(file = "data/boulder-precip.csv")
head(boulder_precip)

ggplot(data = boulder_precip,  aes(x = DATE, y = PRECIP)) +
    geom_point(alpha = 0.9, aes(color = PRECIP)) +
    labs(x = "Date",
      y = "Precipitation (Inches)",
      title = "Daily Precipitation (inches)",
      subtitle = "Boulder, Colorado 2013") + theme_bw(base_size = 9)
```

## Time Series Data
### Lesson 1 - Dates
```{r}
boulder_precip$DATE <- as.Date(boulder_precip$DATE,
                        format = "%m/%d/%y")
class(boulder_precip$DATE)
head(boulder_precip$DATE)
ggplot(data = boulder_precip, aes(x = DATE, y = PRECIP)) +
      geom_bar(stat = "identity", fill = "purple") +
      labs(title = "Total daily precipitation in Boulder, Colorado",
           subtitle = "Fall 2013",
           x = "Date", y = "Daily Precipitation (Inches)")
```
### Lesson 2 - Use Tidyverse Pipes to subset
```{r}
# library(lubridate)
# library(dplyr)
options(stringsAsFactors = FALSE)

# import the data
boulder_daily_precip <- read.csv("data/week-02/precipitation/805325-precip-dailysum-2003-2013.csv",
         header = TRUE,
         na.strings = "999.99")
head(boulder_daily_precip)
str(boulder_daily_precip)
summary(boulder_daily_precip$DAILY_PRECIP)
max(boulder_daily_precip$DAILY_PRECIP)

# Using Pipe
boulder_daily_precip %>%
  mutate(DATE = as.Date(boulder_daily_precip$DATE, format = "%m/%d/%y")) %>%
  na.omit() %>%
  ggplot(aes(x = DATE, y = DAILY_PRECIP)) +
      geom_point(stat = "identity", color = "darkorchid4") +
      labs(title = "Daily precipitation in Boulder, Colorado",
           subtitle = "2003 - 2013",
           x = "Date", y = "Daily Precipitation (Inches)") +
     theme_bw(base_size = 11)

# Subset the data
# subset 2 months around flood
boulder_daily_precip %>%
  mutate(DATE = as.Date(boulder_daily_precip$DATE, format = "%m/%d/%y")) %>%
  filter(DATE >= as.Date("2013-08-15") & DATE <= as.Date("2013-10-15")) %>% 
  ggplot(aes(DATE,DAILY_PRECIP)) +
  geom_bar(stat = "identity", fill = "darkorchid4") +
  xlab("Date") +
  ylab("Precipitation (inches)") +
  xlim(as.Date("2013-08-15"), as.Date("2013-10-15")) +
  ggtitle("Daily Total Precipitation Aug - Oct 2013 for Boulder Creek") +
  theme_bw()

precip_2012 <- boulder_daily_precip %>%
  mutate(DATE = as.Date(boulder_daily_precip$DATE, format = "%m/%d/%y")) %>%
  filter(DATE >= as.Date('2012-08-15') & DATE <= as.Date('2012-10-15'))

ggplot(data=precip_2012, aes(DATE,DAILY_PRECIP)) +
    geom_bar(stat = "identity", fill = "darkorchid4") +
    xlab("Date") +
    ylab("Precipitation (inches)") +
    xlim(as.Date('2012-08-15'), as.Date('2012-10-15')) +
    ylim(0,10) +
    ggtitle("Daily Total Precipitation Aug - Oct 2013 for Boulder Creek") +
    theme_bw()

```

### Lesson 3 - Summarize time series data using Tidyverse pipes
```{r}
boulder_daily_precip %>%
  filter(JULIAN >= 230 & JULIAN <= 290) %>% 
  ggplot(aes(x = JULIAN, y = DAILY_PRECIP)) +
      geom_point(color = "darkorchid4") +
      facet_wrap( ~ YEAR, ncol = 3) +
      labs(title = "Daily Precipitation - Boulder, Colorado",
           subtitle = "Data plotted by year",
           y = "Daily Precipitation (inches)",
           x = "Day of Year") +
      theme_bw(base_size = 15)

# add a month column to your boulder_daily_precip data.frame
boulder_daily_precip <- boulder_daily_precip %>%
  mutate(DATE = as.Date(boulder_daily_precip$DATE, format = "%m/%d/%y")) %>%
  mutate(month = month(DATE))

# calculate the sum precipitation for each month
boulder_daily_precip_month <- boulder_daily_precip %>%
  group_by(month, YEAR) %>%
  summarise(sum_precip = sum(DAILY_PRECIP))

# subset 2 months around flood
boulder_daily_precip_month %>%
  mutate(month2 = as.Date(paste0("2015-", month,"-01"),"%Y-%m-%d")) %>%
  na.omit() %>% 
  ggplot(aes(x = month2, y = sum_precip)) +
      geom_bar(stat = "identity", fill = "darkorchid4") +
      facet_wrap(~ YEAR, ncol = 3) +
      labs(title = "Montly Total Daily Precipitation - Boulder, Colorado",
           subtitle = "Data plotted by year",
           y = "Daily precipitation (inches)",
           x = "Month") + theme_bw(base_size = 15) +
      scale_x_date(date_labels = "%b")

```

### Lesson 4 - HW Challenge: Plot USGS Stream DIscharge Data
```{r}
discharge <- read.csv("data/week-02/discharge/06730200-discharge-daily-1986-2013.csv",
                      header = TRUE)
head(discharge)

discharge %>% 
  mutate(date = as.Date(datetime, '%m/%d/%y')) %>%
  filter(date >= as.Date('2013-08-15')& date <= as.Date('2013-10-15')) %>% 
  ggplot(aes(x = date, y = disValue)) +
  geom_point(stat = 'identity', color = 'purple') +
  labs(title = "Discharge - Boulder, Colorado",
           subtitle = "1986 - 2016",
           y = "Discharge (cfs)",
           x = "Date") +
  theme_bw(base_size = 15)

```
### Lesson 5 - Subset & Aggregate Time Series Precipitation Data
```{r}
precip_boulder <- read.csv("data/week-02/precipitation/805333-precip-daily-1948-2013.csv",
                           header = TRUE,
                           na.strings = 999.99)
head(precip_boulder)
str(precip_boulder)
max(precip_boulder$HPCP)

hist(precip_boulder$HPCP,
     main = "This looks better after the reimporting with\n no data values specified",
     xlab = "Precip (inches)", ylab = "Frequency",
     col = "darkorchid4")

print("how many NA values are there?")
sum(is.na(precip_boulder))

# convert to date/time and retain as a new field
precip_boulder$DATE <- as.POSIXct(precip_boulder$DATE,
                                  format = "%Y%m%d %H:%M")
str(precip_boulder)
nrow(precip_boulder)
daily_sum_precip <- precip_boulder %>%
  mutate(day = as.Date(DATE, format = '%Y-%m-%d')) %>%
  group_by(day) %>%
  summarise(total_precip = sum(HPCP)) %>%
  na.omit()

daily_sum_precip %>%
  filter(day >= as.Date("2013-01-01") & day <= as.Date("2013-10-15")) %>% 
  ggplot(aes(x = day, y = total_precip)) +
  geom_point(stat = 'identity', color = 'purple') +
  labs(
    title = "Hourly Precipitation - Boulder, Colorado",
    subtitle = "1948 - 2013",
    y = "Precipitation (Inches)",
    x = "Date"
  ) +
  theme_bw(base_size = 15)

tail(daily_sum_precip)
nrow(daily_sum_precip)
names(daily_sum_precip)

```

### Lesson 6 - Create Interactive Plots
```{r}
options(stringsAsFactors = FALSE)
discharge_time <- read.csv("data/week-02/discharge/06730200-discharge-daily-1986-2013.csv")

# fix date using pipes
discharge_time <- discharge_time %>%
  mutate(datetime = as.Date(datetime, format = "%m/%d/%y"))
annual_precip <- ggplot(discharge_time, aes(x = datetime, y = disValue)) +
  geom_point() +
  labs(x = "Time",
       y = "discharge value",
       title = "my title")

annual_precip

# Time Series - Plotly
# create interactive plotly plot
ggplotly(annual_precip)

# create time series object
discharge_timeSeries <- xts(x = discharge_time$disValue,
                            order.by = discharge_time$datetime)
# create a basic interactive element
interact_time <- dygraph(discharge_timeSeries)
interact_time

# create a basic interactive element
interact_time2 <- dygraph(discharge_timeSeries) %>% dyRangeSelector()
interact_time2
```




